---
title: Data prep for *Impact of Increasing Ischemic Time on Heart Transplantation Outcome*
author: "Julia (Gege) Ran, William Parker"
output:
  rmdformats::material:
    highlight: tango
  html_document:
    theme: cosmo
    toc: yes
---

# Loading Packages

This chunk is used to load in the packages in R that help us to tidy, manipulate, and visualize the data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

require("knitr")
opts_knit$set(root.dir = "C:/Users/julia/SRP/S&D/SharedFiles/pubsaf2103")

```

```{r library}
library(tidyverse)
library(haven)
library(rmdformats)
library(dplyr)
library("ggpubr")
```

# Data sources

Standard Scientific Registry of Transplant Recipients (SRTR) SAF files We uploaded all the relevant files that contained data for the study.

```{r data_in}
# read in the SRTR SAF files
##all thoracic transplant candidate information
cand_thor <- haven::read_sas("cand_thor.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#justification for status 1a
statjust_hr1a <- read_sas("statjust_hr1a.sas7bdat", NULL) %>%
  zap_formats() %>% zap_labels()

#justification for status 1b
statjust_hr1b <- read_sas("statjust_hr1b.sas7bdat", NULL) %>%
  zap_formats() %>% zap_labels()

#transplant result
tx_hr <- read_sas("tx_hr.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#transplant follow-up result
txf_hr <- read_sas("txf_hr.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#center-level and opo-level predictors 
institution <- read_sas("institution.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

opo <- read_sas("hist_opo_txc.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()

donor <- read_sas("donor_deceased.sas7bdat",NULL) %>% 
  zap_formats() %>%  zap_labels()

```

```{r new justification}
#justification for new statuses
setwd("C:/Users/julia/SRP/S&D/Thoracic/ThoracicRegistration/Thoracic_Justs")

##testtest

#justification for new status 1
statjust_hr1 <- read_sas("JustFormHRStat1.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

#justification for new status 2
statjust_hr2 <- read_sas("JustFormHRStat2.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()

#justification for new status 3
statjust_hr3 <- read_sas("JustFormHRStat3.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()

#justification for new status 4
statjust_hr4 <- read_sas("JustFormHRStat4.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()


#status justification and episodes
StatusJustEpisode <- read_sas("StatusJustEpisode.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()

#linking WlregAuditID to justification ID
JustFormHRDataLink <- read_sas("JustFormHRDataLink.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels()

#linking WlregAuditID to Px_ID, the linking is copied from RiskStratDataHR
WlregAuditId_PxId <- read_sas("RiskStratDataHR.sas7bdat", NULL) %>% 
  zap_formats() %>% zap_labels() %>% select(WlregAuditId,PX_ID=px_id)

#JustFormHR
JustFormHR <- read_sas("JustFormHR.sas7bdat")

#column descriptions
ColumnDescriptions <- read_sas("ColumnDescriptions.sas7bdat")
```

# Select study sample

## 1. Define data ranges

This section specifics the dates used in the study. The dates can be modified by changing the date in this chunk of code.

```{r date_ranges}
start_date <- as.Date("2016-10-20")
end_date <- as.Date("2019-12-31")

#mark the end of the pre-policy cohort
pre_policy_end_date <- as.Date("2017-12-31")


#mark the start of the post-policy cohort
post_policy_start_date <- as.Date("2018-10-20")

policy_switch_date <- as.Date("2018-10-18")
```

## 2. Filter initial listing

In this section, we select only adult, heart-only transplant recipients from sufficiently large transplant centers. The following exclusion criteria were applied:

1.  Age \<= 17 yo at the time of listing
2.  Recipient of multiple transplants
3.  Inactive listing status
4.  Transplant center conducted fewer than 5 heart-alone transplants in either of the two study periods
5.  Did not receive a transplant
6.  Had inappropriate listing status at the time of transplantation (e.g. had Status 1A/1B/2 assignment after the policy switch)

As a result, 6775 transplant recipients from 99 centers are included in further analyses.

```{r filter initial listing}
#keep multi-organ recipients to code as Status 5
multi <- TRUE

#exclude tranplant recipients <18 at the time of listing
peds <- FALSE

init_list <- tx_hr %>% 
  mutate(tx_date = REC_TX_DT) %>% 
  filter(tx_date >= start_date & tx_date <= end_date)  %>% 
  mutate(policy = case_when(
  tx_date <= pre_policy_end_date ~ paste0("Pre"),
  tx_date >= post_policy_start_date ~  paste0("Post"),
  tx_date >= pre_policy_end_date & tx_date <= post_policy_start_date ~ paste0("in-between"))) %>% 
  mutate(tx = ifelse(is.na(REC_TX_DT)== TRUE, 1,0)) %>% 
  filter(tx !=1, policy !="in-between") %>% #exclude candidates w/o transplant or tx-ed between outside the range of study periods
  mutate(center = REC_CTR_ID) %>% 
  filter(CAN_AGE_AT_LISTING >17) %>% #exclude peds
  filter(!(policy == "Post" & (CAN_LAST_STAT == 1010 | CAN_LAST_STAT == 2010)))

init_list %>% group_by(policy) %>% count()

#remove multiorgan recipients
if (multi == FALSE){
	multi_recips <- tx_hr %>% filter(REC_TX_TY == 2 | REC_TX_TY ==4) %>% select(PX_ID,REC_TX_TY)
  
	n_mults <- nrow(init_list %>% filter(PX_ID %in% multi_recips$PX_ID))
	
	init_list <- init_list %>% filter(!PX_ID %in% multi_recips$PX_ID)
	remove(multi_recips)
}

#filter out inactive listing 
init_list <- init_list %>%
  filter(CAN_LAST_STAT %in% 
           c(1010,1020,1030,1110,1120,1130,1140,1150,1160,
             2010,2020,2030,2110,2120,2130,2140,2150,2160)) 


#remove candidates transplanted at low volume centers
init_list <- init_list %>% 
  mutate(transplant = ifelse(CAN_REM_CD == 4, 1, 0)) %>%
	group_by(REC_CTR_ID,policy) %>% 
  mutate(tot_tx = replace_na(sum(transplant, na.rm = TRUE), 0)) %>% 
  ungroup() 

low_volume_centers <- init_list %>% 
  select(REC_CTR_ID , policy,tot_tx) %>% 
  group_by(REC_CTR_ID , policy) %>% 
  distinct() %>% 
  ungroup() %>% 
  spread(key = policy, value = tot_tx) %>% 
  mutate(Pre = replace_na(Pre, 0),
         Post = replace_na(Post,0)) %>% 
  filter(Pre <5 | Post < 5) %>% 
  select(REC_CTR_ID)

low_volume_center <- pull(low_volume_centers, REC_CTR_ID)

init_list <- init_list %>% 
  mutate(low_volume = ifelse(REC_CTR_ID  %in% low_volume_center, 1,0)) %>% 
  filter(low_volume !=1) 


#sanity check - 99 centers are in the study
n_centers <- init_list %>% 
  select(REC_CTR_ID , policy,tot_tx) %>% 
  group_by(REC_CTR_ID , policy) %>% 
  distinct() %>% 
  ungroup() %>% 
  spread(key = policy, value = tot_tx) %>% 
  mutate(Pre = replace_na(Pre, 0),
         Post = replace_na(Post,0)) %>% 
  nrow()

#3367 transplants pre-policy, 3581 post-policy. 
num_tx_period <- init_list %>% 
  select(PX_ID, policy, tx_date, CAN_LAST_STAT) %>% 
  group_by(policy) %>% 
  count()


```

# Identify Predictors of Post-Transplantation Mortality

Next, we identified variables most likely to impact an organ recipient's post-transplantation survival based on components of the IMPACT score:

1.  Age \> 60 years
2.  Serum bilirubin, mg/dl
3.  Dialysis between listing and transplant
4.  Female sex
5.  Heart failure etiology
6.  Recent infection
7.  Intra-aortic balloon pump
8.  Mechanical ventilation pre-transplant
9.  Race
10. Temporary circulatory support (ECMO, VAD)
11. Ventricular assist device

Other explanatory variables under consideration include 

12. Recipient functional status 
13. Recipient and donor ABO blood type 
14. Recipient bmi 
15. Donor age, 
16. Treatment received by the recipient immediately before transplantation 
17. Recipient educational background
18. CAN_CARDIAC_OUTPUT 
19. REC_PCW_MEAN



Additionally, we will identify the following outcome variables

1.  death within 1-year post-transplantation (all causes)
2.  re-transplantation within 1-year post-transplantation
3.  date of graft failure (fatal and non-fatal)

```{r identify variables}

#with IMPACT score components, functional status, and ABO
init_list2 <- init_list %>% 
  select(PX_ID,
         policy,
         tx_date,
         status = CAN_LAST_STAT,
         center = REC_CTR_ID,
         rec_age = REC_AGE_IN_MONTHS_AT_TX,
         don_age = DON_AGE_IN_MONTHS,
         rec_edu = CAN_EDUCATION,
         rec_bmi = REC_BMI,
         bilirubin = REC_TOT_BILI,
         dialysis = REC_DIAL,
         sex = CAN_GENDER,
         diagnosis = CAN_DGN,
         infection = REC_INFECT_IV_DRUG,
         iabp = REC_IABP,
         vent = REC_VENTILATOR,
         no_inhaled = REC_INHALED_NO,
         inotrope = REC_INOTROP,
         pge = REC_PGE,
         life_support = REC_LIFE_SUPPORT,
         other_life_support = REC_LIFE_SUPPORT_OTHER,
         defib = REC_IMPLANT_DEFIB,
         infection_iv = REC_INFECT_IV_DRUG,
         life_support_type = REC_LIFE_SUPPORT_TY,
         vent_event = REC_VENTILATOR_SUPPORT,
         vent_timeframe = REC_VENTILATOR_TIMEFRAME,
         race = CAN_RACE,
         ecmo = REC_ECMO,
         vad = REC_VAD_TY,
         vad_b1 = REC_VAD1,
         vad_b2 = REC_VAD2,
         vad_tah = REC_VAD_TAH,
         rec_abo = CAN_ABO,
         don_abo = DON_ABO,
         fnt_status = REC_FUNCTN_STAT,
         last_stat = TFL_LASTATUS, 
         last_date = TFL_LAFUDATE,
         cod = TFL_COD, 
         gf_date = TFL_GRAFT_DT,
         death_date = TFL_DEATH_DT,
         ischemic_time = REC_HR_ISCH) %>% 
  mutate(
    race_desc = case_when(
      race == 2000 ~ "Hispanic/Latino",
      race == 1024 ~ "Unknown",
      race == 512 ~ "Indian Sub-continent",
      race == 256 ~ "Arab or Middle Eastern",
      race == 128 ~ "Native Hawaiian or Other Pacific Islander",
      race == 64 ~ "Asian",
      race == 32 ~ "American Indian or Alaska Native",
      race == 16 ~ "Black or African American",
      race == 8 ~ "White",
      TRUE ~ "Multi-Racial"
    ),
    rec_age = rec_age/12,
    don_age = don_age/12,
    rec_age_60 = ifelse(rec_age >60, 1,0),
    graft_failed = ifelse(!is.na(gf_date),1,0),
         gf_1y = ifelse(!is.na(gf_date) & 
                          (as.Date(gf_date) - as.Date(tx_date) <365),
                        1,0),
         dead = ifelse(!is.na(death_date) & last_stat == "D", 
                       1,0),
         dead_1y = ifelse(!is.na(death_date) & 
                            (as.Date(death_date) - as.Date(tx_date) <365),
                          1,0),
         retx = ifelse(last_stat == "R",1,0)
  )

#sanity check 1: 160 pre-policy vs 110 post-policy recipients had graft failure post transplant
View(init_list2 %>% 
  group_by(policy,graft_failed) %>% 
  count() %>% 
  filter(graft_failed ==1))

#sanity check 2: 118 pre-policy vs 72 post-policy recipients died of graft failure
View(init_list2 %>% 
       filter(graft_failed ==1 & dead == 1 & gf_date == death_date) %>% 
       group_by(policy) %>% 
       count())

#sanity check 3: 283 pre-policy vs 308 post-policy recipients died within the 1st year due to all causes
View(init_list2 %>% 
       filter(dead_1y == 1) %>% 
       group_by(policy) %>% 
       count())

#sanity check 4: average # of days between date of transplantation and date of graft failure: 367 pre-policy vs 142 post-policy
View(init_list2 %>% 
       filter(graft_failed ==1) %>% 
       mutate(gf_duration = as.Date(gf_date) - as.Date(tx_date)) %>% 
       group_by(policy) %>% 
       summarise(mean = mean(gf_duration)))

#sanity check 5: 8 pre-policy vs 13 post-policy recipients got re-transplanted after graft failure
View(init_list2 %>% 
       filter(graft_failed ==1 & last_stat == "R") %>% 
       group_by(policy) %>% 
       count())

#sanity check 6: average ischemic time is significantly longer after policy change 
ggboxplot(init_list2, x = "policy", y = "ischemic_time", 
          color = "policy", palette = c("#00AFBB", "#E7B800"),
        ylab = "ischemic time", xlab = "policy period")

t.test(ischemic_time ~ policy, data = init_list2, var.equal = TRUE)

#sanity check 7: 8 pre-policy vs 13 post-policy re-transplant 

test1 <- init_list2 %>% 
  select(policy, dead_1y) %>% 
  filter(dead_1y != 0) %>% 
  group_by(policy,dead_1y) %>% 
  count() %>% 
  ungroup()
View(test1)

test2 <- init_list2 %>% 
  select(policy,last_stat) %>% 
  group_by(policy,last_stat) %>% 
  count() %>% 
  ungroup() 

test2$policy <- factor(test2$policy, levels = c("Pre","Post"))


ggplot(test2, aes(last_stat,n, fill=policy)) +
  geom_bar(stat = "identity", position = 'dodge') +
  scale_x_discrete(labels=c("A" = "Alive", 
                            "D" = "Dead",
                            "L" = "Lost to FU",
                            "R" = "Retransplant",
                            "N" = "Not seen"
                            ))

```

## Find recipient treatment at the time of transplantation

First, we will find the treatment each pre-policy candidate received at the time of transplant.


```{r stat_1a}
## Status 1A -> Status 1-4
#select pre-policy listings
status_1a <- init_list2 %>% 
  filter(policy == "Pre" & status == 2010)

just_1a <- statjust_hr1a %>%
  filter(PX_ID %in% status_1a$PX_ID) %>% 
  group_by(PX_ID) %>% 
  mutate(chg_dt_seq = dense_rank(CANHX_CHG_DT)) %>% 
  filter(chg_dt_seq == max(chg_dt_seq)) %>% 
  ungroup() 
  

#remove redundant or erroneous justifications
just_1a <- distinct(just_1a) %>%
  filter(CANHX_FORM_STAT == 4 | CANHX_FORM_STAT == 8) %>%
  distinct(PX_ID, CANHX_CHG_DT, .keep_all = TRUE)

#select key variables
just_1a <- just_1a %>% 
  dplyr::select(PX_ID, 
         CAN_LISTING_CTR_ID, CANHX_CHG_DT,
    CANHX_STAT_TY, 
    CANHX_FORM_STAT, 
    CANHX_DIALYSIS, 
    CANHX_LAB_SERUM_CREAT, 
    CANHX_ADULT_CRITERIA_A, 
    CANHX_ADULT_CRITERIA_B, 
    CANHX_ADULT_CRITERIA_C, 
    CANHX_ADULT_CRITERIA_D, 
    CANHX_ADULT_CRITERIA_E, 
    CANHX_INTRP_DOBU, 
    CANHX_INTRP_DOPA, 
    CANHX_INTRP_MILRIN,
    CANHX_ADMITTED, 
    CANHX_IABP, 
    CANHX_ECMO, 
    CANHX_LVAD_TYPE, 
    CANHX_VAD, 
    CANHX_TAH, 
    CANHX_RVAD_TYPE,
    CANHX_LAB_BILI, 
    CANHX_HEMO_SBP, 
    CANHX_CARD_OUTPUT, 
    CANHX_HEMO_CI, 
    CANHX_HEMO_INTRP_OBTAINED, 
    CANHX_HEMO_BSA,
    CANHX_HEMO_PCWP,
    CANHX_HEMO_MPAP,
    CANHX_DEV_MALFUNCTN,
    CANHX_DEV_VENT_ARRYTHM, 
    CANHX_PHYS_HR_RHYTH,
    CANHX_DEV_INFECT
    ) %>% 
  left_join(cand_thor %>% 
              dplyr::select(PX_ID, 
                     CAN_PCW_MEAN, 
                     CAN_HGT_CM, 
                     CAN_WGT_KG, 
                     CAN_CARDIAC_OUTPUT, 
                     CAN_CARDIAC_OUTPUT_MEDS
                     ))
#adding back select cand_thor hemodynamics
#durable LVAD list for elective 1A time
#lvad_list <- c(205, 206, 208, 210, 216, 217, 223, 224, 230, 231, 232, 233)
durable_list <- c(205, 206, 208, 210, 216, 217, 223, 224, 230, 231, 232, 233, 
                  305, 306, 313, 316, 319, 325, 402)
#non dischargable VADs
non_discharge <- c(201, 203, 204, 209, 215, 218, 221, 222, 225, 226, 227, 228, 
                   234, 301, 302, 303, 309, 310, 311, 320, 321)

just_1a <- just_1a %>% 
  mutate(status = "Status 1A",
         stat_just = case_when(
            CANHX_ADULT_CRITERIA_A ==1~ "Status 1A (MCS for shock)",
            CANHX_ADULT_CRITERIA_B == 1 ~ "Status 1A (MCS complication)",
            CANHX_ADULT_CRITERIA_C == 1 ~ "Status 1A (Mechanical ventilation)",
            CANHX_ADULT_CRITERIA_D == 1 ~ "Status 1A (High dose inotropes)",
            CANHX_ADULT_CRITERIA_E == 1 ~ "Status 1A (Exception)"),
    ino_ci = as.numeric(CANHX_HEMO_CI),
    ino_pcwp = as.numeric(CANHX_HEMO_PCWP),
    n_inos = ifelse(is.na(CANHX_INTRP_DOPA)==FALSE, 1, 0) + 
      ifelse(is.na(CANHX_INTRP_DOBU)==FALSE, 1, 0) + 
      ifelse(is.na(CANHX_INTRP_MILRIN)==FALSE, 1, 0),
    n_inos = ifelse(status != 2010, NA, n_inos),
    multi_ino = ifelse(n_inos > 1, 1, 0),
    single_ino = ifelse(n_inos ==1, 1, 0),
    low_d_ino = ifelse(n_inos> 1 & 
               (CANHX_INTRP_DOPA<3 | CANHX_INTRP_MILRIN <0.25 | CANHX_INTRP_DOBU < 3), 1, 0),
    low_s_ino = ifelse(n_inos ==1 & (CANHX_INTRP_MILRIN <0.5 | CANHX_INTRP_DOBU < 7.5),1,0),
    bsa = 0.007184*(CAN_HGT_CM)^(0.725)*CAN_WGT_KG^(0.425),
    tcr_ci = CAN_CARDIAC_OUTPUT/bsa,
    iabp_no_shock = case_when(
      tcr_ci > 1.8 & CAN_CARDIAC_OUTPUT_MEDS == "N" & (CANHX_IABP==1) ~ 1,
      tcr_ci > 2.0 & (CANHX_IABP==1) ~ 1,
      CAN_PCW_MEAN < 15 & CANHX_IABP ==1 ~ 1,
      is.na(tcr_ci) == FALSE & (CANHX_IABP==1) ~ 0),
    dopa_dose = CANHX_INTRP_DOPA,
    dobu_dose = CANHX_INTRP_DOBU,
    milrin_dose = CANHX_INTRP_MILRIN,
    overtreat = case_when(
      ##Note absence of SBP requirement- patients were on inotropes
        iabp_no_shock == 1 ~ 1,
        CANHX_HEMO_PCWP < 15 & CANHX_ADULT_CRITERIA_D ==1 ~ 1,
        ino_ci>2.2~ 1,
        ino_pcwp <15 ~ 1,
        (ino_ci>1.8 & CANHX_HEMO_INTRP_OBTAINED =="N") ~ 1,
        low_d_ino == 1 | low_s_ino == 1 ~ 1, 
        ino_ci <= 1.8 ~ 0,
        (ino_ci <= 2.2 & CANHX_HEMO_INTRP_OBTAINED =="Y") ~ 0,
        TRUE ~ 0),
    elective_1A = case_when(
      CANHX_ADULT_CRITERIA_A ==1 & CANHX_IABP ==0 & CANHX_ECMO ==0 & CANHX_LVAD_TYPE %in% durable_list ~1,
      TRUE ~ 0),
    Justification = case_when(
      CANHX_ADULT_CRITERIA_E ==1  ~ "Status 3- Exception",
      CANHX_ADULT_CRITERIA_C == 1 ~ "Status 3- Vent",
      CANHX_DEV_VENT_ARRYTHM == 1 & CANHX_ADULT_CRITERIA_B ==1 ~ "Status 1-MSCD with Life Threatening Arrhythmia",
    CANHX_ECMO == 1 ~ "Status 1-VA ECMO",
    CANHX_ADULT_CRITERIA_A ==1 & CANHX_VAD == 1 & is.na(CANHX_RVAD_TYPE)== FALSE & (CANHX_LVAD_TYPE %in% non_discharge)~ "Status 1-Non-Dischargeable BiVAD",
    CANHX_VAD == 1 & CANHX_LVAD_TYPE %in% non_discharge ~ "Status 2-Non-Dischargeable LVAD",
    CANHX_DEV_MALFUNCTN == 1 ~ "Status 2-MSCD Device with Malfunction",
    CANHX_ADULT_CRITERIA_A ==1 & CANHX_IABP ==0 & elective_1A==0  ~ "Status 2-Other MSCD",
    CANHX_IABP ==1 & overtreat == 0 ~ "Status 2-Intra Aortic Balloon Pump",
    CANHX_PHYS_HR_RHYTH == 7 ~ "Status 2-Ventricular Tachycardia (VT)",
    CANHX_PHYS_HR_RHYTH == 8 ~ "Status 2-Ventricular Fibrillation (VF)",
    CANHX_ADULT_CRITERIA_B == 1 ~ "Status 3-MSCD Complication",
    elective_1A == 1 ~ "Status 3-Dischargeable LVAD on elective time",
    overtreat == 0 ~ "Status 3-Multiple inotropes or a single high dose inotrope and hemodynamic monitoring",
    overtreat == 1 & CANHX_IABP == 1~ "Status 3-IABP Downgraded Due to Cardiogenic Shock Requirement",
    overtreat == 1 & CANHX_ECMO == 1~ "Status 3- Downgraded Due to Cardiogenic Shock Requirement",
    overtreat == 1 ~ "Status 4-Inotropes Downgraded Due to Cardiogenic Shock Requirement"
))

just_1a <- just_1a %>%
  mutate(six_status = case_when(
    grepl("Status 1", Justification) ~ "Status 1",
    grepl("Status 2", Justification) ~ "Status 2",
    grepl("Status 3", Justification) ~ "Status 3",
    grepl("Status 4", Justification) ~ "Status 4"
    ))

just_1a <- just_1a %>%
  mutate(treatment = case_when(
    stat_just == "Status 1A (Exception)" | stat_just == "Status 1A (Mechanical ventilation)" ~ "Exception",
    Justification %in% c("Status 1-MSCD with Life Threatening Arrhythmia",
                         "Status 2-MSCD Device with Malfunction",
                         "Status 3-Dischargeable LVAD on elective time",
                         "Status 3-MSCD Complication") ~ "LVAD",
    Justification %in% c("Status 1-Non-Dischargeable BiVAD",
                         "Status 2-Non-Dischargeable LVAD",
                         "Status 2-Other MSCD")~"Other MCS",
    Justification == "Status 1-VA ECMO"~"ECMO",
    Justification %in% c("Status 2-Intra Aortic Balloon Pump",
                         "Status 3-IABP Downgraded Due to Cardiogenic Shock Requirement")~"IABP",
    Justification %in% c("Status 2-Ventricular Tachycardia (VT)",
                         "Status 2-Ventricular Fibrillation (VF)")~ "Exception",
    Justification == "Status 3-Multiple inotropes or a single high dose inotrope and hemodynamic monitoring"~"High-dose Inotropes",
    Justification == "Status 3- Exception" ~ "Exception",
    Justification == "Status 4-Inotropes Downgraded Due to Cardiogenic Shock Requirement"~"High-dose Inotropes")) %>%
  dplyr::select(PX_ID, 
         status, 
         CANHX_CHG_DT, 
         six_status, 
         Justification, 
         treatment, 
         CANHX_HEMO_SBP, 
         CANHX_HEMO_CI, 
         CANHX_HEMO_PCWP, 
         CANHX_HEMO_INTRP_OBTAINED) %>%
  group_by(PX_ID) %>% 
  arrange(PX_ID, CANHX_CHG_DT) %>%
  filter(row_number() ==1)
just_1a %>% group_by(treatment) %>% count(Justification)
```

```{r stat_1b}
status_1b <- init_list2 %>% 
  filter(policy == "Pre" & status == 2020)

just_1b <- statjust_hr1b %>% 
  filter(PX_ID %in% status_1b$PX_ID) %>%
  group_by(PX_ID) %>% 
  mutate(chg_dt_seq = dense_rank(CANHX_CHG_DT)) %>% 
  filter(chg_dt_seq == max(chg_dt_seq)) %>% 
  left_join(cand_thor %>% dplyr::select(PX_ID, CAN_DGN, CAN_PCW_MEAN, CAN_HGT_CM, CAN_WGT_KG, CAN_CARDIAC_OUTPUT)) #adding back select cand_thor hemodynamics and diagnosis codes
  

#remove redundant or erroneous justifications
just_1b <- distinct(just_1b) %>%
  filter(CANHX_FORM_STAT == 4 | CANHX_FORM_STAT == 8) %>%
  distinct(PX_ID, CANHX_CHG_DT, .keep_all = TRUE)

#restrictive, amyloid, CHD, and HCOM, cardiomyopathy diagnoses to upgrade old Status 2 candidates
stat4_diagnoses <- c(1050, 1051, 1052, 1053, 1054, 1099, 1100, 1101, 1102, 
                     1103, 1104, 1105, 1106, 1199, 1200, 1201, 1203, 1205, 1206, 1207, 1208)

#simple recode for Status 1B...
just_1b <- just_1b %>%
  mutate(
    bsa = 0.007184*(CAN_HGT_CM)^(0.725)*CAN_WGT_KG^(0.425),
    tcr_ci = CAN_CARDIAC_OUTPUT/bsa,
    Justification = case_when(
      CANHX_VAD == 1 ~"Status 4-Dischargeable LVAD without discretionary 30 days",
      tcr_ci >= 2.2 & CANHX_CONT_IV_INOTROP == 1 ~ "Status 6-Low dose inotropes but cardiac index too high",
      CAN_PCW_MEAN <= 15 & CANHX_CONT_IV_INOTROP == 1~ "Status 6-Low dose inotropes but wedge too low",
      CANHX_CONT_IV_INOTROP == 1 ~ "Status 4-Inotropes without Hemodynamic Monitoring",
      CAN_DGN %in% stat4_diagnoses ~ "Status 4-Medical diagnoses that qualify for status 4",
    TRUE ~ "Status 4-Exception"))

just_1b <- just_1b %>%
 mutate(status = "Status 1B",
        treatment = case_when(
   Justification %in% c("Status 4-Inotropes without Hemodynamic Monitoring", "Status 6-Low dose inotropes but cardiac index too high", "Status 6-Low dose inotropes but wedge too low") ~"Low-dose Inotropes",
   Justification == "Status 4-Dischargeable LVAD without discretionary 30 days"~"LVAD",
   Justification == "Status 4-Medical diagnoses that qualify for status 4" ~ "None",
   Justification ==  "Status 4-Exception" ~ "Exception")) %>%
  select(PX_ID, status, CANHX_CHG_DT, Justification, treatment, tcr_ci) %>%
  group_by(PX_ID) %>% 
  arrange(PX_ID, CANHX_CHG_DT) %>%
  filter(row_number() ==1)

just_1b <- just_1b %>%
  mutate(six_status = case_when(
    grepl("Status 6", Justification) ~ "Status 6",
    grepl("Status 4", Justification) ~ "Status 4"
    ))

#sanity checks:
just_1b %>% group_by(treatment) %>% count(Justification)

just_1b %>% 
  group_by(Justification) %>%
  summarise(min_CI = min(tcr_ci, na.rm = TRUE))
```

## Combine status 1A and 1B justification data with initial listings from `cand_thor`, code old Status 2 -\> Status 5/6

```{r combine_data_sets}
#restrictive, amyloid, CHD, and HCOM, cardiomyopathy diagnoses to upgrade old Status 2 candidates
stat4_diagnoses <- c(1050, 1051, 1052, 1053, 1054, 1099, 1100, 1101, 1102, 
                     1103, 1104, 1105, 1106, 1199, 1200, 1201, 1203, 1205, 1206, 1207, 1208)

just_1a_1b <- bind_rows(just_1a, just_1b)

just_1a_1b <- just_1a_1b %>% 
  rename(Status = status)

multi_recips <- tx_hr %>% 
  filter(REC_TX_TY == 2 | REC_TX_TY ==4) %>% select(PX_ID,REC_TX_DT)

old_policy_lists <- init_list2 %>%
  filter(policy == "Pre" & status < 2100) %>%
  left_join(just_1a_1b , by = c("PX_ID")) %>%
  mutate(
    Justification = case_when(
      status == 2030 & diagnosis %in% stat4_diagnoses ~ "Status 4-Medical diagnoses that qualify for status 4",
    status == 2030 ~ "Status 6-Adult Candidates Suitable for Transplant",
    TRUE ~ Justification
    ),
    treatment = ifelse(status ==2030, "None", treatment),
    six_status = case_when(
      status ==2030 & diagnosis %in% stat4_diagnoses ~"Status 4", 
      status ==2030 & PX_ID %in% multi_recips$PX_ID ~ "Status 5",
      status ==2030 ~ "Status 6",
      TRUE ~ six_status
      )) %>% 
  mutate(Status = ifelse(status == 2030, "Status 2", Status)) %>% 
  select(-status) %>% 
  rename(status = Status) %>% 
  select(PX_ID,status, six_status, treatment,Justification, contains("CANHX"))
  
old_policy_lists %>% group_by(status) %>% count(treatment)

old_policy_lists %>%  group_by(status) %>% count()
```

Next, we will find the treatment each post-policy candidate received at the time of transplant.

```{r identify post_policy candidates}
post_policy <- init_list2 %>%
  filter(policy == "Post") %>%
  select(PX_ID, 
         tx_date, 
         status)

status_ep_with_PX_ID <- StatusJustEpisode %>%
  left_join(JustFormHRDataLink %>% select(WlregAuditId, JustId), by = "JustId") %>%
  left_join(WlregAuditId_PxId, by = "WlregAuditId") %>%
  filter(PX_ID %in% post_policy$PX_ID) %>%
  left_join(JustFormHR %>% select(JustId, status = RequestedCandStatCd, Exception)) %>%
  select(PX_ID, JustId, status, Exception, ChangeDate) 


post_policy1 <- post_policy %>% 
  left_join(status_ep_with_PX_ID %>% select(-status), by = c("PX_ID")) %>% 
  group_by(PX_ID) %>% 
  mutate(just_sequence = dense_rank(ChangeDate)) %>%
  filter(just_sequence == max(just_sequence)) %>% 
  filter(JustId == max(JustId)) %>% 
  ungroup() 

#for every patient, 1) find the latest change date, 2) if there are duplicates, query all JustId, usually only one of them should be filled

```

## Status 1 form cleaning
```{r status_1}
post_policy_status1 <- post_policy1 %>% 
  filter(status == 2110) %>% 
  pull(JustId)
# select only justifications that appear in first_ep
#coding life-threatening arrythmia as an "exception"

#there should be 311 people transplanted at Status 1
status_1 <- statjust_hr1 %>%
  filter(JustId %in% post_policy_status1) %>%  
  left_join(post_policy1 %>% 
              select(PX_ID, JustId, status, Exception), by = "JustId") %>% 
  mutate(
    treatment = case_when(
      CriteriaEcmoSupport == TRUE ~ "ECMO",
      CriteriaBivadSupport == TRUE ~ "Other MCS",
      CriteriaMcsdSupport == TRUE | Exception == TRUE ~ "Exception" 
    )
  ) %>% select(PX_ID, JustId, treatment, status, Exception, EcmoCardiacIndex, CardiacIndexInotropeSupport, EcmoCapWedgePressure, EcmoWithoutHemo)
status_1 %>% count(treatment)

```

## Status 2 form cleaning
```{r status_2}
#only true durable dischargable LVAD crtieria for Status 2 is malfunction
#labelling VT/VF as an exception
post_policy_status2 <- post_policy1 %>% 
  filter(status == 2120) %>% 
  pull(JustId)

status_2 <- statjust_hr2 %>%
  filter(JustId %in% post_policy_status2) %>%  
  left_join(post_policy1 %>% 
              select(PX_ID, JustId, status, Exception), by = "JustId") %>%
  mutate(
    treatment = case_when(
      CriteriaIabpSupport == TRUE ~ "IABP",
      CriteriaMcsdMalfunction == TRUE ~ "LVAD", 
      CriteriaDurableDevSupport == TRUE |  CriteriaMcsdEndovasSupp == TRUE | CriteriaLvadSupport == TRUE ~ "Other MCS",
      CriteriaVentEpisode == TRUE ~ "Exception", 
      Exception == TRUE ~ "Exception"
    )
  ) %>% select(PX_ID,JustId, treatment, status, IabpCardiacIndex, IabpCapWedgePressure, IabpWithHemo, IabpWithoutHemo, IabpCardiacIndexInotropeSup)
status_2 %>% count(treatment)
status_2 %>% count()
```


## Status 3 form cleaning

### Variables needed for the hemodynamic graphs from high dose inotrope candidates selected

```{r status_3}
post_policy_status3 <- post_policy1 %>% 
  filter(status == 2130) %>% 
  pull(JustId)

status_3 <- statjust_hr3 %>%
  filter(JustId %in% post_policy_status3) %>%
  left_join(post_policy1 %>% 
              select(PX_ID, JustId, status, Exception), by = "JustId") %>%
  mutate(
    treatment = case_when(
      CriteriaPercuSupport == TRUE ~ "Other MCS",
      CriteriaLvadDiscSupport == TRUE | CriteriaMcsdWithHemo == TRUE | CriteriaMcsdWithPump == TRUE | CriteriaMcsdWithRhf == TRUE ~ "LVAD",
      CriteriaMcsdInfection == TRUE | CriteriaMcsdMucosalBleed == TRUE |  CriteriaMcsdWithAI == TRUE | CriteriaLvadSupport == TRUE ~ "LVAD",
      CriteriaInotropeSupport == TRUE ~ "High-dose Inotropes",
      CriteriaVaEcmoSupport == TRUE ~ "ECMO",
      CriteriaIabpSupport == TRUE ~ "IABP",
      Exception == TRUE ~ "Exception"
    )
  ) %>% select(PX_ID,JustId, treatment, status, InoSysBloodPressure, InoCardiacIndex, InoCapWedgePressure, InoInotropeSupport)
status_3 %>% count(treatment)
status_3 %>% count()
```


## Status 4 form cleaning

```{r status_4}
post_policy_status4 <- post_policy1 %>% 
  filter(status == 2140) %>% 
  pull(JustId)

status_4 <- statjust_hr4 %>%
  filter(JustId %in% post_policy_status4) %>%
  left_join(post_policy1 %>% 
              select(PX_ID, JustId, status, Exception), by = "JustId") %>%
  mutate(
    treatment = case_when(
      CriteriaLvadSupport == TRUE ~ "LVAD",
      CriteriaInotropeSupport == TRUE ~ "Low-dose Inotropes",
      CriteriaHeartDisease == TRUE | CriteriaIschemicHeart == TRUE | CriteriaCardiomyopathy == TRUE | CriteriaRetransplant == TRUE ~ "None",
      Exception == TRUE ~ "Exception"
    )
  ) %>% select(PX_ID,JustId, treatment, status, InotropeCardiacIndex, InotropePcwp)
status_4 %>% count(treatment)
status_4 %>% count()
```

## Status 5-6 coding
```{r status_5_6}
post_policy_status56 <- post_policy1 %>% 
  filter(status == 2150 | status == 2160) %>% 
  pull(JustId)


status_5_6 <- post_policy1 %>%
  filter(status %in% c(2150, 2160, 1150, 1160)) %>%
  mutate(treatment = "None" ) %>%
  select(PX_ID,JustId, treatment, status)


status_5_6 %>% count()
```

## Combine and create full post dataset

```{r post-dataset, warning = FALSE, message= FALSE}
init_list2 <- init_list2 %>% 
    mutate(PX_ID = ifelse(is.na(PX_ID)==FALSE,as.integer(PX_ID),PX_ID))

post_justifications <- bind_rows(status_1, status_2, status_3, status_4, status_5_6) %>%
  mutate(status = case_when(
           status %in% c(1110, 2110) ~ "Status 1",
           status %in% c(1120, 2120) ~ "Status 2",
           status %in% c(1130, 2130) ~ "Status 3",
           status %in% c(1140, 2140) ~ "Status 4",
           status %in% c(1150, 2150) ~ "Status 5",
           status %in% c(1160, 2160) ~ "Status 6"
         ),
         six_status = status) 

post_justifications %>% group_by(status) %>% count(treatment)
post_justifications %>% group_by(status)
```


# Final Sample - Without Transplant Center Characteristics

This dataframe contains all transplants matched with their variables used for the analysis.

```{r create_final_sample}
#put together pre and post policy information contained in old_policy_lists and post_justifications respectively


final_sample <- bind_rows(old_policy_lists, 
                          post_justifications %>% select(PX_ID, 
                                                         status, 
                                                         treatment,
                                                         six_status
                                                         )) %>% 
  select(PX_ID, status, treatment, six_status) %>%
  left_join(init_list2 %>% select(-status), by = "PX_ID") 

#fix coding names

final_sample <- final_sample %>% 
  mutate(vad = case_when(
    vad == 1 ~ "None",
    vad == 2 ~ "LVAD",
    vad == 3 ~ "RVAD", 
    vad == 4 ~ "TAH", 
    vad == 5 ~ "LVAD+RVAD",
    vad == 6 ~ "Unspecified"
  ))

final_sample %>% filter(tx_date > as.Date("2019-01-08") & treatment == "Low-dose Inotropes")

final_sample %>% group_by(policy) %>% count(treatment)



#people with contradictory IABP record - n = 103
problem_iabp <- final_sample %>% 
  filter(treatment == "IABP" & iabp == "0") %>% 
  select(PX_ID, status, treatment, policy, life_support, life_support_type, vad, ecmo, no_inhaled,pge,inotrope)

#people with contradictory VAD record - n = 8
problem_vad <- final_sample %>% 
  filter(treatment == "LVAD" & vad == "None") %>% 
  select(PX_ID, status, treatment, policy, life_support, life_support_type, iabp, ecmo)

#people with contradictory ECMO record - n=9
problem_ecmo <- final_sample %>% 
  filter(treatment == "ECMO" & ecmo == "0") %>% 
  select(PX_ID, status, treatment, ecmo, policy, life_support, life_support_type, iabp, vad)

#can we further code vad into older generation, newer generation, or heart mate 2?
```

## Basic sanity check

1. Refine final sample to exclude people missing ischemic time
```{r exclude missing ischemic time}
final_sample1 <- final_sample %>% 
  filter(!is.na(ischemic_time)) 

n_missing_ischemic <- count(final_sample) - count(final_sample1)
#excluded 39 individuals for missing ischemic time
#total n= 6736

sample_size<- final_sample1 %>% group_by(policy, six_status) %>% count()
```

1. Calculate mean ischemic time by status tier, treatment, and policy period
```{r ischemic time by status tier}
#by status
ischemic_by_status <- final_sample %>% #n = 6,737
  filter(!is.na(ischemic_time)) %>% 
  group_by(policy, six_status) %>% 
  summarize(avg=mean(ischemic_time), 
            sd=sd(ischemic_time),
            n=n(), 
            se=sd/sqrt(n)          
)

View(ischemic_by_status)
ggplot(ischemic_by_status, aes(x=six_status, y=avg, group=policy, color=policy)) + 
  geom_pointrange(aes(ymin= avg-sd, ymax=avg+sd))

#by treatment
ischemic_by_treatment <- final_sample %>% #n = 6,737
  filter(!is.na(ischemic_time)) %>% 
  group_by(policy, treatment) %>% 
  summarize(avg=mean(ischemic_time), 
            sd=sd(ischemic_time),
            n=n(), 
            se=sd/sqrt(n)          
)
View(ischemic_by_treatment)
ggplot(ischemic_by_treatment, aes(x=treatment, y=avg, group=policy, color=policy)) + 
  geom_pointrange(aes(ymin= avg-sd, ymax=avg+sd))


#by policy
ischemic_by_policy <- final_sample %>% #n = 6,737
  filter(!is.na(ischemic_time)) %>% 
  group_by(policy) %>% 
  summarize(avg=mean(ischemic_time), 
            sd=sd(ischemic_time),
            n=n(), 
            se=sd/sqrt(n)          
)
View(ischemic_by_policy)
ggplot(ischemic_by_policy, aes(x=policy, y=avg)) + 
  geom_pointrange(aes(ymin= avg-sd, ymax=avg+sd))

t.test(ischemic_time ~ policy, data = final_sample1, var.equal = TRUE)
```

3. calculate survival time for each person
