---
title: Tidyverse Practice
author: "Julia (Gege) Ran, William Parker"
output:
  rmdformats::material:
    highlight: tango
  html_document:
    theme: cosmo
    toc: yes
---

This is a R markdown file to help you get familiar with tidyverse using heart transplant data. By the end of this meeting, you should be able to use the following commands to wrangle a dataset

-   load appropriate packages
-   import data
-   create dataframe
-   filter
-   select
-   mutate
-   join dataframes
-   generate basic summary data
-   generate basic summary graphs
-   export dataset

1. load packages

```{r load packages}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

require("knitr")
opts_knit$set(root.dir = "C:/Users/julia/SRP/S&D/SharedFiles/pubsaf2103") #change this to wherever your data files are stored 

#use install.packages() to download the following packages

library(tidyverse) #so that you don't have do code in Base R
library(haven) #for reading SAS
library(rmdformats) #for knitting your rmd into html
library(dplyr) #dplyr lives in tidyverse

#you will also need ggplor and ggpubr for graphing down the line
```

2. Load tx_hr and cand_thor into your environment

```{r load data}
#you can also specify setwd() in the beginning of this chunk if the data is stored in a different folder than what you have specified in root.dir() 

#setwd("C:/Users/julia/SRP/S&D/SharedFiles/pubsaf2103")
#if your root.dir() and setwd() are different, you have to run setwd() and the code to extract data simultaneously for R to find it. 


cand_thor <- haven::read_sas("cand_thor.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()

tx_hr <- read_sas("tx_hr.sas7bdat", NULL) %>%  
  zap_formats() %>% zap_labels()
```

3.  Define your study time frame

    In this study, we are only interested in patients who received transplant between 2016-10-20 to 2017-12-31 or 2018-10-20 to 2019-12-31.

```{r study time frame}
#pre-policy start and stop time
start_date <- as.Date("2016-10-20")
pre_policy_end_date <- as.Date("2017-12-31")

#post-policy start and stop time
post_policy_start_date <- as.Date("2018-10-20")
end_date <- as.Date("2019-12-31")

#policy switch date
policy_switch_date <- as.Date("2018-10-18")

```

4.  Select your study population

From tx_hr, we select only qualifying adult, heart-only transplant recipients listed in the aforementioned time range.

```{r initial candidate list}
#init_list will be a subset of observations in tx_hr

#1. Using tx_hr, make a dataframe called init_list with a new column called tx_date (transplant date), which is equal in value as the variable REC_TX_DT in tx_hr
init_list <- tx_hr %>% 
  mutate(tx_date = REC_TX_DT)

#2. Using filter(), keep only recipients whose tx_date is within the pre-defined study period and remove everyone who never got a transplant or was transplanted outside of the study period
  
#case_when() - specify the input into a new column based on conditions, when the condition is not boolian

#ifelse() - specify the input into a new column based on boolian conditions

#is.na() - determine if an element is empty (TRUE) or not (FALSE)

#!is.na() - opposite of is.na(), determine if an element is NOT empty 

#"!" in front of most operations means negation

init_list <- init_list %>% 
  filter(tx_date >= start_date & tx_date <= end_date)  %>% 
  mutate(policy = case_when(
    tx_date <= pre_policy_end_date ~ paste0("Pre"),
    tx_date >= post_policy_start_date ~  paste0("Post"),
    tx_date >= pre_policy_end_date & tx_date <= post_policy_start_date ~ paste0("in-between"))) %>% 
  mutate(tx = ifelse(is.na(REC_TX_DT)== TRUE, 1,0)) %>% 
  filter(tx !=1, policy !="in-between")  #exclude candidates w/o transplant or tx-ed between outside the range of study periods.
```

Using mutate() and filter(), we can apply a wide range of inclusion and exclusion criteria, such as:

-   Age \<= 17 yo at the time of listing
-   Recipient of multiple transplants
-   Inactive listing status
-   Transplant center conducted fewer than 5 heart-alone transplants in either of the two study periods
-   Did not receive a transplant
-   Had inappropriate listing status at the time of transplantation (e.g. had Status 1A/1B/2 assignment after the policy switch)

As an exercise, let's try to identify low-volume transplant centers, and drop all observations from these centers. To do that, we need to 

1) find out the total number of heart transplants done by each transplant center in each policy period

2) generate a roster of transplant centers that are low-volume (i.e. # transplant <5 per policy period)

3) drop all observations from these identified transplant centers from init_list


```{r drop low-volume centers step 1}
#find transplant volume per center per policy period

#SRTR data dictionary: CAN_REM_CD describes reasons why a candidate was removed from the waitlist. CAN_REM_CD ==4 means the person was removed because they got a transplant

#create a new binary variable transplant that takes the value of 1 if the person received a transplant, and 0 if the patient was removed due to some other reason. 
init_list2 <- init_list %>% 
  mutate(transplant = ifelse(CAN_REM_CD == 4, 1, 0)) %>%
	group_by(REC_CTR_ID,policy) %>% 
  mutate(tot_tx = replace_na(sum(transplant, na.rm = TRUE), 0)) %>% 
  ungroup() 

#select() - keep only the specified columns
#group_by() - carry out the following commands for each group, until ungroup() is specified
#distinct() - keep only distinct rows
#spread() - reshapes the dataset from long to wide by values in a key variable
low_volume_centers <- init_list2 %>% 
  select(REC_CTR_ID , policy,tot_tx) %>% 
  group_by(REC_CTR_ID , policy) %>% 
  distinct() %>% 
  ungroup() %>% 
  spread(key = policy, value = tot_tx) %>% 
  mutate(Pre = replace_na(Pre, 0),
         Post = replace_na(Post,0)) %>% 
  filter(Pre <5 | Post < 5) %>% 
  select(REC_CTR_ID)

#alternatively, what happens if we use the count() function?
#count() - sums up the number of rows in the whole dataset or by group 
low_volume_center2 <- init_list %>%
  filter(CAN_REM_CD ==4) %>% 
  group_by(REC_CTR_ID, policy) %>% 
  count() %>% 
  ungroup() %>% 
  select(-policy) %>% 
  filter(n<5) %>% 
  select(-n) %>% 
  distinct() 
  
# we end up finding fewer low-volume centers because count() will not include rows where n=0. So if center #112 had 12 transplants pre-policy and 0 post-policy, count() won't make a row for the group (center #112 + post_policy). As a result, we would miss this center.   
```

```{r drop low-volume centers step 2}
#generate a roster of transplant centers that are low-volume (i.e. # transplant <5 per policy period)

#pull() - converts a datafram to a vector
low_volume_center <- pull(low_volume_centers, REC_CTR_ID)

```

```{r drop low-volume cetners step 3}
#drop all observations from these identified transplant centers from init_list

# %in% allows you to filter observations by their presence in another vector/dataframe
init_list2 <- init_list2 %>% 
  mutate(low_volume = ifelse(REC_CTR_ID  %in% low_volume_center, 1,0)) %>%
  filter(low_volume !=1) 
```

5.  Keep only candidate ID, policy period, name of the center that they are transplanted at, and the date of their transplant from init_list2.
```{r simplify init_list2}
init_list3 <- init_list2 %>% 
  select(PX_ID, policy, tx_date, REC_CTR_ID)

```

6. Find the diagnosis, mean PCW, height, weight and cardiac output of each patient in cand_thor and join that column to init_list3 to create init_list4

```{r join cand_thor}
#left_join() - use dataframe before the chain as a base and join columns from another dataset on the left side. 

#by: specifies which row from the outside dataset should be joined with which row in the present dataset
init_list4 <- init_list3 %>% 
  left_join(cand_thor %>% dplyr::select(PX_ID, CAN_DGN, CAN_PCW_MEAN, CAN_HGT_CM, CAN_WGT_KG, CAN_CARDIAC_OUTPUT), by = "PX_ID")

```

7. Do some sanity check. What is the size of each policy cohort? 

```{r sanity check}
# see how I am not using the assign "<-" symbol? I don't need to save this result, I just want to see it. 
init_list4 %>% 
  group_by(policy) %>% 
  count()

#to see it in a different window, put the same commands in View(). Remember to capitalize V.
View(init_list4 %>% 
  group_by(policy) %>% 
  count())
```
